---
title: "R Notebook"
output: html_notebook
---

Question 5: Basic R Programming for Big Data (20 Points)


a)
```{r}
# load bigmemory library
library(bigmemory)
# open ratings data file in big matrix format
ratings <- read.big.matrix(filename = "ratings.dat", sep=",", type = "integer",
                           col.names = c("UserID", "ProfileID", "Rating"))
```

```{r}
# calculate the weighted rank of each profile id used by IMDB. That is, for v = number of votes for profile, R=average rate for profile, m = the 250th largest number of ratings for a single profile (4182), C = mean over whole data, calculate (v/(v+m))*R+(m/(v+m))*C
weighted.rank <- function(ProfileID) {
  # get the indexes of all the ratings which have the same profileids
  ratingIndexes <- mwhich(ratings, "ProfileID", ProfileID, "eq")
  ratingIndexes
  # get profile ratings of unique profiles
  allProfileRatings <- ratings[ratingIndexes, ]
  allProfileRatings
  # calculate mean of profile ratings
  R <- mean(allProfileRatings[,3])
  # calculate get the number of votes for the given profile
  v <- nrow(allProfileRatings)
  # m
  m <- 4182
  # get mean of whole data
  C <- mean(ratings[,3])
  # calculate weighted rank
  weightedrank <- (v/(v+m))*R+(m/(v+m))*C
  # return the weighted rank
  return(weightedrank)
}

```

```{r}
# get all profiles rated by user 100
profilerates <- mwhich(ratings, "UserID", 100, "eq")
allProfileRates <- ratings[profilerates, ]
weightedprofiles <- lapply(allProfileRates[,2], weighted.rank)
weightedprofiles <- unlist(weightedprofiles)

```

```{r}
# plot histogram of the scores calculated
library(ggplot2)
ggplot() + geom_histogram(aes(weightedprofiles), bins = 5)
```



```{r}
# Lets find ratings of male users coming from New York and ratings of female users coming from California.
load("users.Rdata") # load data
# the data we have opened contains many variants of each state. In order to capture the full essence of the data, we should see all variants of States New York and California. We do this by grabbing all unique state names which begin with n or c
grep("n", unique(User[,3]), value = TRUE)
unique(User[,2], value=TRUE)
grep("c", unique(User[,3]), value = TRUE)
```

```{r}
# create dataset of males and new york
newyorkm <- User[User$Gender %in% c("M") & User$State %in% c("New York", "new york", "nY", "ny"), ]
# create dataset of females and california
calif <- User[User$Gender %in% c("F") & User$State %in% c("ca", "california", "calif.", "caifornia", "ca.", "cA"), ]
#get ratings of males coming from New York
nym <- ratings[ratings[,1] %in% newyorkm$UserID, ]
#get ratings of females coming from California
calf <- ratings[ratings[,1] %in% calif$UserID, ]
nym <- data.frame(location_gender = "New York Males", rating = nym[, 3])
calf <- data.frame(location_gender = "California Females", rating = calf[, 3])
# bind datasets
nycaliratings <- rbind(nym, calf)
# plot boxplot comparing data
ggplot(nycaliratings, aes(x=location_gender, y=rating)) + geom_boxplot()
```

Spread is similar, there are no outliers and it appears ratings of females coming from California have a higher rating than ratings of males which are coming from New York. 


Now lets begin attempting to predict user's rating for a profile. We seek to create a model for predicting user a's rating on profile b. This will be done using a Multiple Linear Model of 
prediction = x1 + x2(Avg rating given by user a) + x3(Avg rating given to profile b) + error

```{r}
# This subsection is intellectual property of Qiang Sun
X <- ratings
N=3000000 # number of rating records
Nu=135359 # maximum of UserID
Np=220970 # maximum of ProfileID
user.rat=rep(0,Nu) # user.rat[i] denotes the sum of ratings given by user i
user.num=rep(0,Nu) # user.num[i] denotes the number of ratings given by user i
profile.rat=rep(0,Np) # profile.rat[i] denotes the sum of ratings given to profile i
profile.num=rep(0,Np) # user.rat[i] denotes the number of ratings given to profile i
# In each iteration, we update the four arrays, i.e. user.rat, user.num, profile.rat and profile.num, using one rating record.
for (i in 1:N) { 
  user.rat[X[i,'UserID']]=user.rat[X[i,'UserID']]+X[i,'Rating'] # The matrix X   here comes from the file ’ratings.dat’
  user.num[X[i,'UserID']]=user.num[X[i,'UserID']]+1
  profile.rat[X[i,'ProfileID']]=profile.rat[X[i,'ProfileID']]+X[i,'Rating']
  profile.num[X[i,'ProfileID']]=profile.num[X[i,'ProfileID']]+1 
  if (i %% 10000==0) print(i/10000)
}
user.ave=user.rat/user.num #calculate user average
profile.ave=profile.rat/profile.num #calculate profile average
X1=X #compile data into X1 matrix
```
```{r}
head(X)
```


```{r}
options(bigmemory.allow.dimnames=TRUE)
colnames(X1)=c('UsrAveRat','PrfAveRat','Rat')
a <- X[, 1]
X1[,"UsrAveRat"] = user.ave[X[, 1]]
X1[,'PrfAveRat']=profile.ave[X[,2]] # X1 is the new data matrix we will work with in regression.

```

Compute the model based on our dataset
```{r}
library(biganalytics)
head(X1)
#test <- X1[1:20,]
linreg.biglm <- biglm.big.matrix(Rat ~ UsrAveRat + PrfAveRat, data = X1)
coefficients(linreg.biglm)
summary(linreg.biglm)$rsq
```

predict any user's rating

```{r}
curdata <- read.big.matrix(filename = "ratings.dat", sep=",", type = "integer",
                           col.names = c("UserID", "ProfileID", "Rating"))
head(curdata)
```
```{r}
unique(curdata[,1]) # Displays all possible UserIDs
pred.userId <- readline(prompt="Enter UserID you would like to predict: ") # give UserID from list above in console
unique(curdata[,2]) # Displays all possible ProfileIDs
pred.profileId <- readline(prompt="Enter the profileID of the profile whose rating prediction will be given for your UserId: ") # give ProfileID from list above in console
# convert inputs to integers
pred.userId <- as.integer(pred.userId)
pred.profileId <- as.integer(pred.profileId)
# get average rating for the User
avgUser <- curdata[curdata[,1] == pred.userId,]
avgUser <- mean(avgUser[,3])
# get the average rating given by profile
avgProfile <- curdata[curdata[,2] == pred.profileId,]
avgProfile <- mean(avgProfile[,3])
# combine averages into a data frame
topred <- merge(avgUser,avgProfile)
colnames(topred) <- c("UsrAveRat", "PrfAveRat")

# predict the rating
prediction <- predict(linreg.biglm,newdata = topred)
prediction

# get real value (If no real value exists, this will return 0)
curdata[(curdata[,1] == pred.userId) & (curdata[,2] == pred.profileId),3]

```
